# Appendix {-}

## Git and Session Info

This code lists the versions of R and all the R packages I'm using, tells me which computer and operating system processed these files, and connects this output to its git commit.
```{r cache=FALSE}
#don't cache this otherwise it won't update
print(system("hostname",intern=T))
print(system("git log -1",intern=T)[1:3])
print(sessionInfo(),locale=F)
#save sessioninfo to file
writeLines(capture.output(sessionInfo()), "sessioninfo.txt")
```

## Save Workspace

Save workspace 
so data can be retrieved at later time.  Normally not something I do but given data are changing so often might as well keep a snapshot of the data according to each release. I may decide to stop saving the data.  File name will change when I change the version number on git (e.g., v0.9).

```{r}
#version was defined at the beginning of the the index.Rmd file
#get current tag and use a filename prefix
currentversion <- system("git describe --tags",intern=T)
save.image(file=paste0("data-archive/",currentversion,"-dataworkspace.RData"))
print(paste0("File name associated with this run is ", currentversion, "-dataworkspace.RData"))
```
