# Deeper Understanding {#deeperunderstanding}

In this chapter I consider some hypotheses that can be tested with relatively simple methods.

## Regression Discontinuity

This subsection will show a fun test that we can examine in real time.  Some governors of US states have decided to reopen businesses much sooner than recommended by public health officials.  For example, Georgia has decided to reopen gyms, massage therapy, tatoo shops, barbers and hair stylists on April 24th with restaurants, theaters and private social clubs opening up on the 27th.  It will be  difficult to practice social distancing while getting a tatoo or a body piercing.

Pending:  I'll plot the data for Georgia by county 3 weeks before the 24th and 3 weeks after.  It will take a few days to see the effects on the positive testing rate due to a higher rate of social contact, and maybe up to two weeks to see any effects on the death rate.  I'll examine a similar plot for an adjacent state that has not relaxed their social distancing mandate for comparison.

This example will illustrate the use of the regression discontinuity (RD) design, though before the casuality police get on my case for not clarifying my assumptions I'll point out that I won't be using this to claim I have estimated the causal effect of social distancing.  I'll illustrate the idea of this design, its characteristic plot and  point to papers where the reader can learn about the various approaches to extracting additional information from this type of research design.


```{r warning=F, message=F, eval=F, echo=F}
datacov.US.long <- gather(datacov.US, day, count, `1/22/20`:(names(datacov.US)[ncol(datacov.US)]))
datacov.US.long$day <- as.Date(datacov.US.long$day, "%m/%d/%y")
datacov.US.long$day.numeric <- as.numeric(datacov.US.long$day)-18283


georgia.df <- subset(datacov.US.long, `Province/State`=="Georgia")

ggplot(subset(georgia.df, day > as.Date("2020/04/15")), aes(day,count, group=Admin2, color=Admin2)) + geom_line() + ggtitle("Georgia Counties")

```

## Hospital Readmission Data

My colleague Jack Iwashyna who you will learn more about in the concluding chapter, suggested a hypothesis in  mid April 2020.  [Research](https://www.ncbi.nlm.nih.gov/pubmed/28550403) suggests that 40% of  ARDS patients are re-admitted to the hospitial within 1 year of discharge. This estiamte came from a national, multicenter study with 839 ARDS survivors.  Most covid-19 deaths are related to ARDS (more on this in the concluding chapter).   Granted traditional ARDS patients may be different than covid-19-related ARDS patients but let's use this as a plausible estimate of hospital re-admission.  If we will see roughly a 40% chance of hospital readmission, it means that we may see additional hospitalizations in the coming months on top of the already steady stream of new covid-19 hospitalizations. And in the fall, if the predicted "second covid-19 wave" hits that could be compounded by additional hospital readmissions of people recently discharged or currently in the hospital. As the months progress, I'll present analyses examining this hypothesis along with models that will assess second waves.

The good news is that the study found that patient-reported physical activity and quality of life status is associated with fewer re-admissions.  So if we can get these  covid-19-related ARDS survivors walking and exercising we may have some hope. Unfortunately, many ARDS survivors have trouble merely standing up from a chair let alone going for a walk around the block wearing a mask.

More on this to come.


```{r eval=F, echo=F}
ggplot(subset(covid.tracking, date > as.Date("2020/03/20")), aes(date, hospitalizedCurrently, group=state, color=state)) + geom_line()

```

## Deaths

Pending
