# Process Models


Process models are models where we model mechanisms. A use of such models is that they allow us to model "what if" scenarios, such as what would be expected to happen to the spread of covid-19 cases if a state imposes a shelter-in-place order.  The models in the previous chapters don't allow us do such analyses.  The material in the earlier chapters allow us to justify some of the assumptions we may need in process models.  Sometimes process models can be estimated directly from data, sometimes data merely provide ways of justifying assumptions one needs to make.

For a basic example of how we can use basic data along with sum assumptions to estimate the "actual number of cases" check out this [video](https://www.youtube.com/watch?v=mCa0JXEwDEk).

A readable tutorial on process models in understanding pandemics is the book [Charting the Next Pandemic by Pastore y Piontti et al](file:///Users/gonzo/Dropbox/transfer/mac%20transfer/Ana%20Pastore%20y%20Piontti,%20Nicola%20Perra,%20Luca%20Rossi,%20Nicole%20Samay,%20Alessandro%20Vespignani%20-%20Charting%20the%20Next%20Pandemic_%20Modeling%20Infectious%20Disease%20Spreading%20in%20the%20Data%20Science%20Age-Springer%20International%20.pdf). It was published prior to the current covid-19 pandemic.

## SIR Models

SIR models refer to a modeling approach that uses differential equations, which are equations that govern how a process changes over time.  The basic model starts by having three kinds of people:  Susceptible, Infectious and Recovered (SIR).  In various versions of these models one could change assumptions such as "once an individual recovers they will not get the disease again because they develop antibodies" to "an individual who recovers has a p chance of being reinfected if they come in contact with an infected person" (where p is a probability between 0 and 1 and could be varied to study the implications). The [wiki page](https://en.wikipedia.org/wiki/Compartmental_models_in_epidemiology) on such models is not bad.

[Here](https://mail.google.com/mail/u/0/#inbox/QgrcJHrhwzzjMlbFSsGcwGcnZcVmwGkBHMl?projector=1) is a 20min video on SIR models geared for the pandemic with examples around social distancing and [Washington Post article](https://www.washingtonpost.com/graphics/2020/world/corona-simulator/) using an SIR approach to illustrate the effects of social distancing. The latter page appear below in small interactive window (the cool web page within a web page trick).


```{r echo=F}
knitr::include_url("https://www.washingtonpost.com/graphics/2020/world/corona-simulator/", height="600px")

```

Some UM researchers have started analyzing the covid-19 pandemic using SIR models, these include [Peter Song](https://sph.umich.edu/faculty-profiles/song-peter.html) ([slides1](http://www.umich.edu/~songlab/COVID-19.pdf)), [Qianying Lin](https://midas.umich.edu/faculty-member/qianying-lin/) ([using data from Wuhan](https://www.ijidonline.com/article/S1201-9712(20)30117-X/fulltext), [Marisa Eisenberg](https://sph.umich.edu/faculty-profiles/eisenberg-marisa.html) (prediction dashboard](https://epimath.github.io/covid-19-modeling/#current-forecasts)))

There are several R packages for SIR modeling, including [EpiModel](https://www.epimodel.org/), [tsiR](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0185528)

```{r echo=F}
knitr::include_url("https://www.washingtonpost.com/graphics/2020/world/corona-simulator/", height="600px")

```

<!-- eSIR site: http://www.socr.umich.edu/people/dinov/courses/DSPA_notes/18_BigLongitudinalDataAnalysis_RNN.html

song lab slides https://github.com/lilywang1988/eSIR
-->


## Network-based Models

## Agent-based Models

## Public Policy Models

## Summary

The process models I've reviewed in this chapter have a strong public health and epidemiological perspective. We model the data of yesterday to help us understand what is going on and  evaluate potential changes that could be used to intervene on the spread.

Overall, there is a sense that these models are not very satisfying. They operate at one level of analysis: counts on aggregate geographic areas such as countries and states.  They don't get into levels of specific human behavior nor evaluation of policies, contextual factors, health disparities and other aspects that wouldn't be too difficult to gather even more relevant data and build more complex models.   

But there are also other levels that these notes have been quite silent about---the biological level. The SARS-CoV-2 virus itself has many levels of analysis such as molecular, cellular, and genetic as well as complex interactions that we are just beginning to understand in how the virus interacts with its environment (e.g., how it travels in air, how long it can live on a surface, how it interacts with the host).
[Here](https://www.youtube.com/watch?v=vPtH42Lnt_Y) is a wonderful video on some relevant biology including what the SARS-CoV-2 virus does to the lungs and how in some people it causes [Acute Respitory Distress Syndrome](https://www.thoracic.org/patients/patient-resources/breathing-in-america/resources/chapter-2-acute-respiratory-distress-syndrome.pdf) (ARDS).  ARDS is a common cause of death among individuals with covid-19, which occurs because of the massive inflammatory reaction that covid-19 patients experience. 

I would love to see deeper models that integrate the biological, behavioral, social and epidemiological factors.  No single level of analysis will help us figure out how to best respond to all aspects of the current pandemic.  


